---
title: "Mini-Project"
author: "Harshwardhan Patil, Manasa Gudise, Radhika Agarwal"
date: "2023-03-07"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r  include=FALSE, echo=FALSE}
library(ggplot2)
library(tidyverse)
library(broom)
library(gapminder)
```

```{r  include=FALSE, echo=FALSE}
summary(gapminder)
```


## Question 1: GDP and life expectancy in 2007
##### Note: we will be dropping continent Oceania, as it contains few countries conpared to the other continents.

```{r include=FALSE,echo=FALSE}
data = gapminder %>% subset(continent != "Oceania")
```


```{r include=FALSE,echo=FALSE}
data_year_2007 = data %>% filter(year == '2007')
head(data_year_2007)
```

#### Question 1.1: How does life expectancy vary with GDP per capita in 2007? 
#### Answer: 
To view the relationship between the GDP per Capita and life expectancy, we will plot a scatter plot. Scatter plot is used to describe relationship between two variables.

```{r echo=FALSE}
ggplot(data_year_2007, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  ggtitle("Relationship between GDP Per Capita and Life Expectancy") +
  labs(x = "GDP Per Capita", y = "Life Expectancy", color = "Continent") + 
  theme_light()
```
In 2007, the life expectancy increases with the increase in the GDP per Capita. However, the relationship between GDP Per Capita and Life Expectancy is not linear. By Looking at the plot, it appears that the relationship will be best described by logarithmic transformation of GDP per Capita because for low values of GDP, life expectancy increases rapidly, however for high values of GDP, life expectancy increases with mild progression.


#### Question 1.2: Can the trends be well-described by a simple model such as a linear model, or is a more complicated model required?
#### Answer: 
As we have mentioned in 1.1, the relationship will be best described by logarithmic transformation of GDP per Capita, So we will apply log transformation to `gdpPercap` variable.

```{r echo=FALSE}
ggplot(data_year_2007, aes(x = log10(gdpPercap), y = lifeExp, color = continent)) +
  geom_point() +
  ggtitle("Relationship between Log of GDP Per Capita and Life Expectancy") +
  labs(x = "Log of GDP Per Capita", y = "Life Expectancy", color = "Continent") + 
  theme_light()
```

Looking at the above plot, it might be possible that the newly formed relationship between Log of GDP Per Capita and Life Expectancy can be described by linear model. So, we will try to fit a regression line between the data points.

```{r echo=FALSE}
ggplot(data_year_2007, aes(x = log10(gdpPercap), y = lifeExp)) +
  geom_point(aes(color = continent)) +
  geom_smooth(method="lm", se = FALSE) +
  ggtitle("Relationsip described by using a Linear Model") +
  labs(x = "Log of GDP Per Capita", y = "Life Expectancy", color = "Continent") + 
  theme_bw()
```

As we can see from the above plot, Linear model explains the relationship pretty well. However, from 3.25 to 4.0, it does a poor job in data fitting. So, we have to look at differnet data fitting methods.
One of the method is to fit the data using LOESS smoother.

LOESS smoother is a non parametric method, which means it will take no assumptions about the distribution of the data. 
The basic idea behind LOESS is to fit a weighted regression line to the data, where the weights are higher for points that are closer to the point being estimated and lower for points that are further away. This results in a smooth curve that captures the underlying trend in the data while reducing the impact of random fluctuations or noise.

```{r echo=FALSE}
ggplot(data_year_2007, aes(x = log10(gdpPercap), y = lifeExp)) +
  geom_point(aes(color = continent)) +
  geom_smooth(method="loess", color='red', se = FALSE) +
  ggtitle("Relationsip described by using a Loess Smoother") +
  labs(x = "Log of GDP Per Capita", y = "Life Expectancy", color = "Continent") + 
  theme_light()
```

From the plot, we can conclude that, the LOESS does a reasonably good job than linear model for fitting the data.


#### Question 1.3: Is the pattern the same or different for every continent? If some continents are different, which ones?
#### Answer: 
We can try to fit the LOESS smoother(as mentioned in 1.2), for every continent, to view the relationship in every continent.
```{r echo=FALSE}
ggplot(data_year_2007, aes(x = log10(gdpPercap), y = lifeExp, color = continent)) +
  geom_point() +
  geom_smooth(method="loess", se = FALSE) +
  ggtitle("Scatter Plot with LOESS smoother for each Continent") +
  labs(x = "Log of GDP Per Capita", y = "Life Expectancy", color = "Continent") + 
  theme_light()
```

To plot the relationship individually we can use facet grid for each continent.

```{r fig.width=10, fig.height=4,  echo=FALSE}
ggplot(data_year_2007, aes(x = log10(gdpPercap), y = lifeExp, color = continent )) +
  geom_point() +
  geom_smooth(method = "loess", se=FALSE, method.args = list(degree = 2)) +
  facet_grid(. ~ continent) +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("Relationship with LOESS smoother for each Continent") +
  labs(x = "Log of GDP Per Capita", y = "Life Expectancy", color = "Continent") + 
  theme_bw()  
```

From the above plots, we can infer that the patterns are different for each continent. 
* The relationship trend for Asia and America is almost similar, however it is different for Africa and Europe.
* Countries in Africa generally have a lower GDP per capita and life expectancy than rest of the continents and Europe has the higher GDP per capita.
* The pattern explained by the Africa, America and Asia is fairly linear, but thats not the case with Europe.


#### Question 1.4: Can differences between continents be simply described by an additive or multiplicative shift, or is it more complicated than that? 

As we know, we can determine whether the differences between continents described by an additive or multiplicative shift using a Box plot to compare the distribution.

If the box plots for each continent would have similar shapes but varying center values, then we can implement additive shift and if the the box plots for each continent would have comparable spreads but varying positions on horizontal axis then we can implement a multiplicative shift. 

```{r echo=FALSE}
ggplot(data_year_2007, aes(x = continent, y = lifeExp, fill = continent)) +
  geom_boxplot() +
  scale_fill_discrete(name = "Continent") +
  ggtitle("Life Expectancy by Continent in 2007") +
  labs(x = "", y = "Life Expectancy", color = "Continent")  + 
  theme_bw()  
  
```
From the plot, we can see that neither the spread or shape of box plot looks similar. So it is evident that, the differences between continents are more complicated than a simple additive or multiplicative shift, we may need to use more complex models to describe the relationship between GDP per capita and life expectancy.

#### Question 2: Life expectancy over time by continent:

#### Question 2.1 : Has average life expectancy changed over time in each continent? 
#### Answer: 
In order to visualize the change over time in each continent, first of all we will plot life expectancy vs. the year plot for each continent using facet grid.

```{r fig.width=10, fig.height=4, echo=FALSE}
ggplot(gapminder, aes(x = year, y = lifeExp, color = continent, alpha =0.2 )) +
  geom_point() +
  facet_grid(. ~ continent) +
  ggtitle("Life Expectancy by Continent") +
  labs(x = "Year", y = "Life Expectancy", color = "Continent")  + 
  theme_bw()  +
  theme(axis.text.x = element_text(angle = 90))  
```
As we can see, this plot is distorted as it is showing values of life expectancy for each countries present in the respective continent for every year. The one way to get around this, is to take a mean of the life expectancy for each year per continent.

To accomplish this, we will use `group by` and `summarize` function in R. As the average population in China is greater than other continents we will use weighted average by population. For that we will use `weighted.mean`.

```{r echo=FALSE}
summarised_data= gapminder %>%
  group_by(year, continent) %>%
  summarize(avg_life_expectancy = weighted.mean(lifeExp, w = pop))

```
Now, using `summarised_data` we can visualize the average life expectancy change over time for each continent.

```{r fig.width=10, fig.height=4, echo=FALSE}
ggplot(summarised_data, aes(x = year, y = avg_life_expectancy, color = continent )) +
  geom_point() +
  geom_smooth(method = "loess", se=FALSE, method.args = list(degree = 2)) +
  facet_grid(. ~ continent) +
  ggtitle("Life Expectancy by Continent") +
  labs(x = "Year", y = "Life Expectancy", color = "Continent")  + 
  theme_bw()  +
  theme(axis.text.x = element_text(angle = 90)) 
```

The average life expectancy has almost linearly changed over time in each continent.


#### Question 2.2: Have some continents caught up (at least partially) to others? If so, is this just because of some countries in the continent, or is it more general? 
### Answer: 
Yes. Asia caught up in life expectancy with america. If we have a look at the average life expectancy, in the early part(year 1950) people in America had higher average life expectancy than people in Asia. But this has been changed over time and we can say that Asia managed to increased its life expectancy over time, and thus caught up with America. This is not the case with people in Europe and Africa.
We can see that from plot 2.1, few countries in Africa have caught up with countries in the america and Asia despite having much lower life expectancy in the begining. Also, for few countries in Asia, life expectancy was below than the countries in America and Europe, but at the end, these countries had higher life expectancy than rest of the continent.

To get in-depth understanding of the variation of life expectancy over the years, we can plot a box plot and look at the distribution.
```{r fig.width=8, echo=FALSE}
# create a box plot of life expectancy by continent and year
ggplot(gapminder, aes(x=continent, y=lifeExp, fill=continent)) + 
  geom_boxplot() + 
  facet_wrap(~ year, ncol = 6) + 
  labs(title = "Life Expectancy by Continent and Year", x = "Continent", y = "Life Expectancy")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90)) 

```

Looking at the above box plots, we can surely say that in ternms of life expectency the continent of Asia is cought up with America.
Also, Europe managed to caught up with Oceania till the year 1987. 
This trend is more general and not just due to a few countries within each continent. However, there are still differences within continents, particularly in Asia, Europe and America where some countries have much life expectancy than others.


#### Question 2.3: Have the changes been linear, or has it been faster/slower in some periods for some continents? 
#### Answer:
We can plot a line plot to visualize the relationship.

```{r  echo=FALSE}
ggplot(summarised_data, aes(x = year, y = avg_life_expectancy, color = continent)) +
  geom_point() +
  geom_line() +
  labs(title = "Rate of change of Life Expectancy over the period", x = "Year", y = "Average Life Expectancy", color = "Continent") + 
  theme_bw()
```

Yes, for the most of the years, the changes in the average life expectancy have been linear. However, The plot shows that the rate of change in average life expectancy has not been constant over time and across continents. 
In Africa, a rapid increase in the average life expectancy is observed till year 1987 and after that it become steady for a while and then saw a gradual increase.
For Oceania, the increment in life expectancy was much slower in the initial years, and after 1972, the life expectancy started to climb rapidly.
Whereas, Europe and America shown a more consistent but slower improvement over time. Asia saw a steady incraese, before dropping in 1957. This decrease in life expectancy continued till 1962. From 1962 to 1967, a steep increase in the life expectancy. After 1967, the life expectancy increased gradually.

#### Question 2.4: What might explain periods of faster/slower change?
There could be many factors that explain periods of faster or slower change in average life expectancy.
Some factors like changes in population and it's distribution could be significant. Also, higher GDP per capita often correlates with better access to healthcare and other resources that impact life expectancy. Additionally, technological advancements, political and social factors such as wars, epidemics, and changes in government policies also impact the life expectancy.



#### Question 3: Changes in the relationship between GDP and life expectancy over time:
#### Question 3.1: How has the relationship between GDP and life expectancy changed in each continent?
#### Answer: 
To analyze the relationship between GDP and life expectancy over time, we can create scatter plots of life expectancy versus GDP per capita for each continent and color the points by year.

```{r fig.width = 10, fig.height=4, echo=FALSE}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = year)) +
  geom_point() +
  facet_grid(. ~ continent) +
  geom_smooth(method = "loess", se=FALSE, color = 'red', method.args = list(degree = 2)) +
  labs(title = "Relationship between Life expectancy and GDP per Capita in each Continent", x = "GDP per Capita", y = "Life Expectancy", color = "Year") + 
  theme_bw()
```
We can view this as a variation of year as well.

```{r fig.width = 10, fig.height=4, echo=FALSE}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = factor(year))) +
  geom_point() +
  facet_grid(. ~ continent) + 
  geom_smooth(method = "loess", se=FALSE, color = 'red', method.args = list(degree = 2)) +
  labs(title = "Relationship between Life expectancy and GDP per Capita in each Continent", x = "GDP per Capita", y = "Life Expectancy", color = "Year") + 
  theme_bw()
```


This seems like a logarithmic trend. So we can say that, the relationship between GDP and life expectancy in each continent can be expalined logarithmically. So, we will now transform gdpPerCapita logarithmically, and make it a linear relationship between gdpPerCap and lifeExp.

```{r fig.width = 10, fig.height=4, echo=FALSE}
ggplot(gapminder, aes(x = log10(gdpPercap), y = lifeExp, color = factor(year))) +
  geom_point() +
  facet_grid(. ~ continent) +
  geom_smooth(method = "loess", se=FALSE, color = "red",  method.args = list(degree = 2)) +
  labs(title = "Relationship between Life expectancy and Log GDP per Capita in each Continent", x = "Log GDP per Capita", y = "Life Expectancy", color = "Year") + 
  theme_bw()
```



#### Question 3.2: Can changes in life expectancy be entirely explained by changes in GDP per capita? Does it look like there’s a time effect on life expectancy in addition to a GDP effect?
#### Answer:
In order to explain the changes, we can plot a correlation matrix and findout, whether there is strong correlation exists between life expectancy and other variables, such as year.

```{r}
library(GGally)

ggpairs(gapminder[,c("year", "gdpPercap", "lifeExp", "pop")], title = "Corelation plot between various attributes")+
  theme_bw()

```
 So, the changes in life expectancy cannot be entirely explained by gdpPerCapita. It appears that there is strong correlation exists between life expectancy and year as well. 
  
```{r fig.width = 10, fig.height=4, echo=FALSE}
ggplot(gapminder, aes(x = log10(gdpPercap), y = lifeExp, color = factor(year), alpha=0.001)) +
  geom_point() +
  facet_grid(. ~ continent) +
  geom_smooth(method = "lm", se=FALSE) +
  labs(title = "Relationship between Life expectancy and Log GDP per Capita in each Continent modeled on each year", x = "Log GDP per Capita", y = "Life Expectancy", color = "Year") + 
  theme_bw()
```

For each continent, the slope of the smoother is different which shows that the changes in life expectancy are not similar and there exists a variation which can be explained by other factors such as time as well. As the time increases the life expectancy tends to increase as well, even though there is no sudden rise in the log gdpPerCapita.

#### Question 3.4: Has there been “convergence” in the sense that perhaps GDP and/or continent don’t matter as much as it used to? Are there exceptions to the general patterns?
We can state that there is some evidence of convergence in the relationship between GDP per capita and life expectancy over time. The scatter plots for different continents do show some similarities in terms of the overall trend, although there are still some differences in the strength of the relationship between GDP per capita and life expectancy.

However, there are also some exceptions to the general patterns. For example, some countries in Africa continue to have very low life expectancies despite increases in GDP per capita. In addition, there are some countries in Asia and Europe that have higher life expectancies than we would expect based on their GDP per capita.


```{r fig.width = 10, fig.height=4, echo=FALSE}
ggplot(gapminder, aes(x = log10(gdpPercap), y = lifeExp, color = factor(continent), alpha = 0.2)) +
  geom_point() +
  geom_smooth(method='lm', se = FALSE) +
  facet_grid(. ~ year) +
  labs(title = "Relationship between Life expectancy and Log GDP per Capita in each year modeled on each country", x = "Log GDP per Capita", y = "Life Expectancy", color = "Continent") + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90))
```

Overall, if we look at different years, the slope of loess smoothers is not always similar for each continent over the time period.
Till 1987 there has been no convergence observed. But After 1987 there has been convergence as the slope of the smoothers for continents America, Asia, Europe and Oceania is almost similar, Africa being an exception.






